The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-03 ms/ms.Rmd
<edit ms>

The first component of reference geometry added, and principal assumption, is a reference vector. A sampling ratio of 100 percent was used to apply the reference vector on a revolving axis, after which a reference point was added by projecting it atop the mesh surface at the intersection of the reference vector where it exits the base of the vessel. A reference plane was inserted using the pick multiple points function, by adding a series of 10 points around the circumference of the bottle's base. Each element of reference geometry (vector, point, and plane) was then used in an interactive 3-2-1 alignment where the vessel was aligned to a global origin, orienting it in 3D space where it sits upright atop a planar surface (assumed to be the intent of the maker). Following alignment, the reference plane and point were deleted.

The widest profile is defined as the location on a mesh that lies farthest from the central point where the reference vector exits the vessel base while oriented atop the planar surface. To identify that location, a mesh sketch will be generated with the planar method using the plane at the base of the vessel to identify and sketch the widest vessel circumference. By using the plane located at the base of the vessel for the sketch, the point at which the reference vector exits the mesh remains linked to the remainder of the reference geometry. A circle will then be sketched using the vector as the centre, extending outward until the whole of the vessel fits within. Using the mesh sketch, a cylinder (surface) will be extruded around the vessel. The accuracy analyser in Dx will then be used to identify the point on the vessel with the lowest deviation from the extruded surface, and a plane (MPlane) will be inserted coplanar to the vector and oriented to the widest point, bisecting each vessel at its' widest profile.

Using the MPlane as the basis for a second mesh sketch, a spline with 15 interpolation points will be sketched on one rim. Above that sketch, a horizontal line will be added where both the spline and horizontal line are used to determine the horizontal tangent of the rim. A vertical line was subsequently added that bisected the rim at the location of the tangent. This operation was repeated for the opposing rim. The addition of this added step was necessary because surface scanners are unable to collect data from the interior of the bottles, so the spline needed to be cut in a replicable location. Since some Caddo bottles exhibit slightly inverted-to-vertical rims, the preceding step was extended to include an additional measure. A line was drawn between each rim tangent, then a second from the intersection of the line and reference vector to a point 10 mm down the vector, where a horizontal line (parallel with the rim peaks) was inserted to intersect with both external walls of the bottle. It is at this intersection that the final mesh sketch was cut to discriminate between the neck and rim. While this step admittedly appears odd in the context of a comparison of bottle shapes that all exhibit direct rims, it is of considerable import for inter-type comparisons where other bottle types exhibit differing rim morphologies (i.e., everted, etc.).

Using the MPlane as the basis for a third sketch, a spline will be populated for the entirety of the silhouetted profile. That spline will be split at the location of the horizontal tangents and the remaining sections that continued into the bottle interior were deleted. The second split will be added at the intersection of the spline and reference vector (centre of base). Four additional splits are subsequently added at the juncture of the base/body and body/neck on each side of the vessel at the points of highest curvature. The point of highest curvature used to split the spline is identified using the curvature function in Dx, and does not represent an arbitrary location.

@@

Landmarks and equidistant semilandmarks were exported as x, y, and z coordinate data from Dx. Those data were aligned to a global coordinate system (Kendall 1981, 1984; Slice 2001), achieved through generalized Procrustes superimposition (Rohlf and Slice 1990) performed in R 4.1.2 (R Development Core Team 2022) using the geomorph library v.4.0.1 (Adams et al. 2018; Adams and Otárola-Castillo 2013). Procrustes superimposition translates, scales, and rotates the coordinate data to allow for comparisons among objects (Gower 1975; Rohlf and Slice 1990). The geomorph package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Rohlf 1999; Slice 2001). The mean consensus configuration and Procrustes residuals were calculated using a generalized Procrustes analysis (GPA). This initial view of the data demonstrates the degree of variability in the aggregated sample of Caddo bottles. As an exploratory measure, GM methods—to include GPA—aid in clarifying shape differences as well as the production of novel a posteriori hypotheses (Mitteroecker and Gunz 2009).

@@

Principal components analysis (Jolliffee 2002) was used as an exploratory means of visualizing shape variation among the vessels. The shape changes described by each principal axis are commonly visualized using thin-plate spline warping of a reference 3D mesh (Klingenberg 2013; Sherratt et al. 2014). Principal components analysis (PCA) was conducted on scaled, translated, and rotated landmarks and semilandmarks, and demonstrates that the first two PC’s account for 55 (PC1) and 25 (PC2) percent of the variation in bottle shape, with each remaining PC representing eight or fewer percent of the variation. The first two PCs are plotted in Figure XX, where warp grids represent the shape changes that occur in PC1 and PC2.

@@

A residual randomization permutation procedure (RRPP; n=10,000 permutations) was used for all Procrustes ANOVAs (Adams and Collyer 2015; Collyer and Adams 2018), which has higher statistical power and a greater ability to identify patterns in the data should they be present (Anderson and Ter Braak 2003). To assess whether shape differs by size (allometry) and site, Procrustes ANOVAs (Goodall 1991) were run that enlist effect-sizes (z-scores) computed as standard deviates of the generated sampling distributions (Collyer et al. 2015). A Procrustes ANOVA was also run to assess whether shape changes with size. The assumption of allometric slope homogeneity was tested with the procD.allometry function using the PredLine option (Adams and Nistri 2010). Should this test not yield a significant result, then allometric slopes are similar--if not identical--across the categories used in the analysis.

A Procrustes ANOVA was used to test for allometry. Results indicate that allometry is significant (RRPP=10,000, Rsq=0.07151, Pr(>F) = 0.0256), meaning that vessel shapes vary significantly with vessel size. Plots of predicted allometric trajectories for bottles and jars are presented in Figure XX. The null hypothesis of parallel slopes is not rejected by the homogeneity of slopes test for group allometries; however, it is apparent that larger bottles and jars differ more than the smaller forms. A second Procrustes ANOVA was used to test for a significant difference in bottle/jar shape, which rose to the level of significance (RRPP=10,000, Rsq=0.0974, Pr(>F) = 0.0035); however, the test for a difference in vessel size did not. The tests of morphological disparity did not yield significant results, indicating that the bottles and jars occupy a similar range of morphospace.

>>>>

The first component of reference geometry added was a reference vector. A sampling ratio of 100 percent was used to apply the reference vector on a revolving axis, after which a reference point was added by projecting it atop the mesh surface where it intersects with the reference vector at the point that the vector exits the base of the vessel. A reference plane was inserted using the pick multiple points function, by adding a series of 10 points around the circumference of the bottle's base. Each element of reference geometry (vector, point, and plane) was subsequently used in an interactive 3-2-1 alignment where the vessel was aligned to a global origin, orienting it in 3D space where it sits upright atop a planar surface (assumed to be the intent of the maker). Following alignment, the reference plane and point were deleted.

The widest vessel profile is defined as the location on a mesh that lies farthest from the central point where the reference vector exits the vessel base while oriented atop the planar surface. To identify that location, a mesh sketch was generated with the planar method using the plane at the base of the vessel to identify and sketch the widest vessel circumference. By using the plane located at the base of the vessel for the sketch, the point at which the reference vector exits the mesh remains linked to the remainder of the reference geometry. A circle was then sketched using the vector as the centre, extending outward until the whole of the vessel fit within. Using a mesh sketch, a cylinder (surface) was extruded around the full extent of the vessel. The accuracy analyser in Dx was used to identify the point on the vessel with the lowest deviation from the extruded surface, and a plane (MPlane) was inserted coplanar to the vector and oriented to the widest point, bisecting the vessel at its' widest profile.

Using the MPlane as the basis for a second sketch, a spline was populated for the entirety of the silhouetted profile. That spline was subsequently split at the location of the horizontal tangents and the remaining sections that continued into the bottle interior were deleted. The second split was added at the intersection of the spline and reference vector (centre of base). Four additional splits were then added at the juncture of the base/body and body/neck on each side of the vessel at the points of highest curvature. The point of highest curvature used to split the spline is identified using the curvature function in Dx, and does not represent an arbitrary location.

@@

Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/gpa.csize.png}
\caption{Results of a, generalised Procrustes analysis; and b, boxplot of centroid sizes for Chimu bottles and jars.}
\label{fig:gpa}
\end{figure}

@@

Principal components analysis \citep{RN1746} was used to visualise shape variation among the elliptical bifaces, and the scatterplot and convex hulls represent the dispersion of shapes in tangent space (Figure \ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape ranges described by each principal axis is commonly visualized using thin-plate spline warping of a reference image \citep{RN1731,RN479}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/pca-warp-botjar.jpg}
\caption{PCA summarizing shape variation for Chimu bottles and jars. }
\label{fig:pca}
\end{figure}

@@

To assess whether shape changes with size (allometry), and whether shape and size differed by group (bottles and jars), Procrustes ANOVAs \citep{RN1749} were run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions \citep{RN1756}. A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs \citep{RN1655,RN11775}, which has higher statistical power and a greater ability to identify patterns in the data should they be present \citep{RN1719}.

Vessel shape differs significantly between bottles and jars (RRPP=10,000, Rsq=0.0974, Pr(>F) = 0.0035); however, vessel size does not (Figure \ref{fig:mshape}) ([Supplementary Materials, Ch. 4](https://seldenlab.github.io/peru.bottle.jar/)). A subsequent analysis of bottle neck/rim and body/base morphology demonstrated that while the neck/rim of Chimu bottles and jars differs significantly in shape (RRPP = 10,000, Rsq = 0.27367, Pr(>F) = 0.0005), body/base shape remains consistent (Figure \ref{fig:mshape}) ([Supplementary Materials, Ch. 5](https://seldenlab.github.io/peru.bottle.jar/)).

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/mshape.compare.png}
\caption{Means shapes and comparisons for Chimu bottles and jars, as well as necks and rims. }
\label{fig:mshape}
\end{figure}

### 2022-11-03 ms/ms.tex
<edit ms>

Landmarks and equidistant semilandmarks were exported as x, y, and z
coordinate data from Dx. Those data were aligned to a global coordinate
system (Kendall 1981, 1984; Slice 2001), achieved through generalized
Procrustes superimposition (Rohlf and Slice 1990) performed in R 4.1.2
(R Development Core Team 2022) using the geomorph library v.4.0.1 (Adams
et al.~2018; Adams and Otárola-Castillo 2013). Procrustes
superimposition translates, scales, and rotates the coordinate data to
allow for comparisons among objects (Gower 1975; Rohlf and Slice 1990).
The geomorph package uses a partial Procrustes superimposition that
projects the aligned specimens into tangent space subsequent to
alignment in preparation for the use of multivariate methods that assume
linear space (Rohlf 1999; Slice 2001). The mean consensus configuration
and Procrustes residuals were calculated using a generalized Procrustes
analysis (GPA). This initial view of the data demonstrates the degree of
variability in the aggregated sample of Caddo bottles. As an exploratory
measure, GM methods---to include GPA---aid in clarifying shape
differences as well as the production of novel a posteriori hypotheses
(Mitteroecker and Gunz 2009).

@@

Principal components analysis (Jolliffee 2002) was used as an
exploratory means of visualizing shape variation among the vessels. The
shape changes described by each principal axis are commonly visualized
using thin-plate spline warping of a reference 3D mesh (Klingenberg
2013; Sherratt et al.~2014). Principal components analysis (PCA) was
conducted on scaled, translated, and rotated landmarks and
semilandmarks, and demonstrates that the first two PC's account for 55
(PC1) and 25 (PC2) percent of the variation in bottle shape, with each
remaining PC representing eight or fewer percent of the variation. The
first two PCs are plotted in Figure XX, where warp grids represent the
shape changes that occur in PC1 and PC2.

@@

A residual randomization permutation procedure (RRPP; n=10,000
permutations) was used for all Procrustes ANOVAs (Adams and Collyer
2015; Collyer and Adams 2018), which has higher statistical power and a
greater ability to identify patterns in the data should they be present
(Anderson and Ter Braak 2003). To assess whether shape differs by size
(allometry) and site, Procrustes ANOVAs (Goodall 1991) were run that
enlist effect-sizes (z-scores) computed as standard deviates of the
generated sampling distributions (Collyer et al.~2015). A Procrustes
ANOVA was also run to assess whether shape changes with size. The
assumption of allometric slope homogeneity was tested with the
procD.allometry function using the PredLine option (Adams and Nistri
2010). Should this test not yield a significant result, then allometric
slopes are similar--if not identical--across the categories used in the
analysis.

A Procrustes ANOVA was used to test for allometry. Results indicate that
allometry is significant (RRPP=10,000, Rsq=0.07151, Pr(\textgreater F) =
0.0256), meaning that vessel shapes vary significantly with vessel size.
Plots of predicted allometric trajectories for bottles and jars are
presented in Figure XX. The null hypothesis of parallel slopes is not
rejected by the homogeneity of slopes test for group allometries;
however, it is apparent that larger bottles and jars differ more than
the smaller forms. A second Procrustes ANOVA was used to test for a
significant difference in bottle/jar shape, which rose to the level of
significance (RRPP=10,000, Rsq=0.0974, Pr(\textgreater F) = 0.0035);
however, the test for a difference in vessel size did not. The tests of
morphological disparity did not yield significant results, indicating
that the bottles and jars occupy a similar range of morphospace.

>>>>

Landmark data were aligned to a global coordinate system
\citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through
generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R
using the \texttt{geomorph} and \texttt{RRPP} packages
\citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition
translates and rotates the coordinate data to allow for comparisons
among objects, while also scaling each biface using unit-centroid
size---the square root of the sum of squared distances from each
landmark to the specimen's centroid
\citep{RN11139,RN11140,RN11564,RN478}. The \texttt{geomorph} package
uses a partial Procrustes superimposition that projects the aligned
specimens into tangent space subsequent to alignment in preparation for
the use of multivariate methods that assume linear space (Figure
\ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/gpa.csize.png}
\caption{Results of a, generalised Procrustes analysis; and b, boxplot of centroid sizes for Chimu bottles and jars.}
\label{fig:gpa}
\end{figure}

@@

Principal components analysis \citep{RN1746} was used to visualise shape
variation among the elliptical bifaces, and the scatterplot and convex
hulls represent the dispersion of shapes in tangent space (Figure
\ref{fig:pca}) \citep{RN8633,RN5616,RN11143,RN7550}. The shape ranges
described by each principal axis is commonly visualized using thin-plate
spline warping of a reference image \citep{RN1731,RN479}.

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/pca-warp-botjar.jpg}
\caption{PCA summarizing shape variation for Chimu bottles and jars. }
\label{fig:pca}
\end{figure}

@@

To assess whether shape changes with size (allometry), and whether shape
and size differed by group (bottles and jars), Procrustes ANOVAs
\citep{RN1749} were run that enlist effect-sizes (zscores) computed as
standard deviates of the generated sampling distributions
\citep{RN1756}. A residual randomization permutation procedure (RRPP; n
= 10,000 permutations) was used for all Procrustes ANOVAs
\citep{RN1655,RN11775}, which has higher statistical power and a greater
ability to identify patterns in the data should they be present
\citep{RN1719}.

Vessel shape differs significantly between bottles and jars
(RRPP=10,000, Rsq=0.0974, Pr(\textgreater F) = 0.0035); however, vessel
size does not (Figure \ref{fig:mshape})
(\href{https://seldenlab.github.io/peru.bottle.jar/}{Supplementary
Materials, Ch. 4}). A subsequent analysis of bottle neck/rim and
body/base morphology demonstrated that while the neck/rim of Chimu
bottles and jars differs significantly in shape (RRPP = 10,000, Rsq =
0.27367, Pr(\textgreater F) = 0.0005), body/base shape remains
consistent (Figure \ref{fig:mshape})
(\href{https://seldenlab.github.io/peru.bottle.jar/}{Supplementary
Materials, Ch. 5}).

\begin{figure}\centering
\includegraphics[width=\linewidth]{figs/mshape.compare.png}
\caption{Means shapes and comparisons for Chimu bottles and jars, as well as necks and rims. }
\label{fig:mshape}
\end{figure}
